Verification of OpenSSL's SHA3 \keccak\ function using the Software Analysis Workbench

### 2021-12-01 paper/proof.tex
Add proof overview

SHA 3 Details
\begin{compactitem}
  \item Process 1600 bits at a time: $5 * 5 * 64$
  \item State vector is $5*5*64*$
  \item Incoming message padded out to whole number of 1600 bit bit-strings
  \item Processed in order by \texttt{KECCAK}
  \item Keccak turns each block into a state vector (rearranges bit layout in memory)
  \item Keccak has 24 rounds for state vector. 
  \item There are 5 hash functions sequentially applied in each round to iteratively update the state vector
  \item After the last round, keccak turns the state vector back into the bit-string./\item The next unprocessed block is \emph{sponged} (e.g., absorbed) into the current bit-string from the latest keccak and the process is repeated 
  \item The max digest (hash) size is 512 bits in the standard
  \item The final bit-string from keccak after sponging and processing every block in the original message is truncated to the desired digest size 
\end{compactitem}

SAW Details
\begin{compactitem}
  \item Cryptol is a specification language specifically for defining crypto-primitives
  \item It has a compiler to turn Cryptol to SAW core terms
  \item It has an execution environment to run test vectors on specifications
  \item SAW is a tool to prove equivalence between SAW core terms
  \item It uses symbolic execution and must unroll every loop to a fixed length
  \item It can take as input C and Java code and compile those to core terms
  \item It can use overides to replace C and Java code in the term
\end{compactitem}

SAW provides a verification environment for compositional proofs of functional equivalence between two pieces of supported code (C, Java, or Cryptol).
SAW uses symbolic execution to create computational models of the code and creates problems of functional equivalence between those two models that can be pushed off to an SMT solver.  
The difficulty of these SMT problems is directly affected by the size and complexity of the computations that are being compared.
This can lead to drastic increases in proof times and sometimes even prevents completion as computations increase.  
SAW's solution to this is to create contracts (called overrides) of a functions input to output relationship which allows for a modular approach to writing the proof.
By breaking up computation into smaller pieces, and creating overrides for these pieces, SAW can often simplify an otherwise complex (and thus incompletable) proof to remain within the capacity of an SMT solver.

Cryptol is a domain-specific specification language, which has been used in the previously mentioned works by Galois' and by Decker et. al to write visually reviewable implementations of NIST standards.
It has strong ties to the SAW tool and, in fact, can be compiled directly down to the SAW Terms (computational models) which are used by SAW in proving functional equivalence.
Because of the interwoven nature of Cryptol with SAW, the Terms created from Cryptol are in most cases significantly simpler than any created through symbolic execution of C/Java code.

The published NIST standards provide no compileable implementation that can be used as a starting point for proving out the algorithm.
As a result, any proof of correctness with a NIST standard is limited in strength to how simple it is use visual review to connect code used in the proof to the NIST standard, coupled with a small sampling of test vectors.
As shown in \ref{proofStructure}, The proof flow starts with a specification of the SHA3 algorithm in Cryptol with the goal of being as similar visually to the NIST standard as possible.
This work discusses significant ways that Cryptol's functional nature can be leveraged to aid in this goal.
Along with simplifying the visual inspection process, this Cryptol specification also passes all test vectors provided by NIST.

The end goal of this proof is to show functional equivalence between this visually reviewable specifications Keccak function, and that of OpenSSL's C code.
Unfortunately, a direct proof between this specification, and OpenSSL's implementation proved unfeasible, because OpenSSL varies from the NIST standard in how it stores the bits of the intermediate state array in memory.
These minor modifications in memory storage put such a proof out of SAW's capabilities, because it relies on strict typing, and comparison of a Cryptol functions input/output to a C functions memory state.
In order to bridge that gap, a second Cryptol specification is included in the proof flow.
It matches in every way the first Cryptol code, except when it comes to how the state array is stored and accessed (at which point it follows OpenSSL's form).
Because of the efficient nature of Cryptol's compilation to SAW Terms, proving equivalence between these Cryptol implementations was possible at the top level of the SHA3 algorithm (which allows for the differences in state array storage to be fully abstracted out).
Due to the strict typing nature required for writing SAW proofs, the scope of this equivalence proof is limited to a given input message size.
The only computations that change in the SHA3 algorithm due to message size are in the padding and sponging of a block.
With this in mind, an exhaustive proof run over each input message byte count from 1 to the block size is sufficient to cover the differences in computation.
The artifact repository accompanying this work contains the receipts of this exhaustive proof. (REFERENCE TO PROOF!!!!)

INNER GREEK FUNCTIONS:

CONTRACTS CAN BE USED TO OVERRIDE AND REPLACE C WITH Cryptol

WITH THESE CONTRACTS SAW CAN COMPLETE A PROOF OF EQUIVALENCE BETWEEN TWO PIECES OF CODE.

This work relies on the correctness of several tools, including language compilers.

\noindent \textbf{Trusted Code Base}:
\begin{itemize}
  \item LLVM bitcode compiler
  \item The Software Analysis Workbench (SAW) including the conversion from LLVM bitcode to SAWCore terms, the Cryptol compiler to SAWCore terms, and SAW's interface with SMT solvers.
  \item The z3, abc, and yices SMT solvers.
\end{itemize}

>>>>

\begin{figure*}[t]
  \centering
  \includegraphics[width=\linewidth]{figs/proof.png}

  \caption{Equivalence proof for the \shaThree\ standard and the \openssl\ C implementation.}
  \label{fig:proofStructure}

\end{figure*}

The goal of the work presented in this paper is to prove the \openssl\ implementation of \shaThree\ matches the \fips\ publication by \nist.
\figref{fig:proofStructure} illustrates the proof strategy.
The proof begins on the left of the figure with the box labeled \emph{NIST Standard}. 
This box represents the \fips\ standard that defines the \shaThree\ algorithm with english prose and math.
The standard also provides several test inputs with the correspondingly  expected digests.
The first step of the proof reproduces the \fips\ \shaThree\ description in \cryptol.
That is the box to the right of the \emph{NIST Standard} box labeled \emph{Spec}.

The equivalence between the \fips\ standard and the \cryptol\ specification is argued by visual inspection along with tests over the published test vectors with their expected digests.
\shaThree\ is described algorithmically in the standard making use of universal quantifiers and other looping structures in the definition.
\cryptol\ is functional in that it uses list comprehensions for iteration, and it does not have quantifiers.
Visually certifying a specification of \shaThree\ that uses list comprehensions directly for quantification and looping is difficult at best---the list comprehensions are too complex to read, and they obscure the algorithmic, and mathematic, structure in the published standard.

The syntactic disconnect between the quantification and looping in the published standard and the list comprehensions in the \cryptol\ specification is bridged with a library that hides the list comprehensions with appropriately named functions.
The final specification with the function hiding the list comprehensions has a one-to-one correspondence to the published standard that is easy to visually certify.
The specification is further validated by tests in \cryptol\ over the published test vectors.
The digests computed from the \cryptol\ specification exactly match those published in the standard on each given input.
The equivalence arrow between the \emph{NIST Standard} box and the \emph{Spec} box represents the visual certification and test vector results.
\secref{sec:fips} details the specification and library.

The extracted core term from the \openssl\ C implementation of \shaThree\ is too big and complex for \saw\ to reason about directly.
Adding to the complexity of the equivalence proof is that the \openssl\ implementation changes the meaning of the state $S$ in the algorithm so that it no longer directly matches the meaning defined in the published standard.
The change effectively reorders the matrix definition of the state to be more amenable to memory and aspects of the \keccak\ implementation in C.
The change is significant enough to make it extremely difficult to argue manually that the computation implemented in \openssl\ matches that defined in the standard.

The next step of the equivalence proof creates a \cryptol\ specification that matches the state reordering in the \openssl\ implementation in an effort to show an equivalence between it and the \cryptol\ specification for the standard.
This new specification is the box in the right middle of \figref{fig:proofStructure} labeled \emph{OpenSSL} with the \emph{CRY} annotation in the upper right corner.
\saw\ is not able to reason symbolically about input and digest sizes to construct a general proof of equivalence between the published standard and the \openssl\ \cryptol specifications, but it is able to prove equivalence between the two specifications given specific sizes for the input and digest.

A series of \saw\ proofs over a range of input sizes with a digest size fixed at $256$-bits is used to prove the equivalence between the published standard and the \openssl\ \cryptol specifications.
The series of proofs varies the input size from zero bits to $1,087$ bits to reflect the rate $1,088$ bit rate required for a 256-bit digest.
The proofs establish the correctness of the input padding and covers the situation where the padding adds an additional $1,088$-bit block to the input 
As such, the meaning of the equivalence arrow between the \emph{Spec} box and the \emph{OpenSSL} box in the middle of the figure is that the two specifications have equivalent computation for a $256$-bit digest for message sizes from $0$ bits to $1,087$ bits.

\begin{figure*}[t]
  \centering
  \includegraphics[width=\linewidth]{figs/proof2.png}

  \caption{Inner Function Contracts}
  \label{fig:proofStructure2}

\end{figure*}

The rest of the equivalence proof between the standard and the \openssl\ implementation centers on the \keccak\ algorithm as shown in the bottom right of \figref{fig:proofStructure}.
The meaning of the equivalence arrow between the \keccak\ definition in the \cryptol\ specification based on the \openssl\ implementation and the actual implemented \keccak\ algorithm in the \openssl\ C implementation is that their computation exactly matches for any input state.
That proof cannot be constructed directly by \saw\ since the extracted core term from the C implementation is too big for symbolic execution when considering all twenty-four rounds required by \keccak.
The full proof requires overrides.

The proof of equivalence between the specification of \keccak\ and the C implementation of \keccak\ in \openssl\ relies on overrides in \saw as shown in \figref{fig:proofStructure2}.
Here, \saw\ proves equivalence between the \cryptol\ specification and the C implementation for each of the five functions that comprise one round of \keccak\ given some arbitrary input state.
That is the meaning of the right hand side of \figref{fig:proofStructure2}.
These are equivalent on any input.

The implementations for each of the five \keccak\ functions in the C implementation are replaced by the corresponding equivalent \cryptol\ specifications in the extracted model from C.
\saw\ is then able to prove the C implementation of \keccak equivalent to the \cryptol\ specification of the \openssl\ implementation of \keccak\ over all twenty-four rounds for any arbitrary input state.
The equivalence is independent of the digest size, message size, or state.

%% TODO: update the figure and text with results from the proof of the padding and the squeeze functions.

In summary, the \cryptol\ specification of the \fips\ standard relies on visual inspection and test vectors for equivalence.
The equivalence between the \cryptol\ specification of the standard and the \cryptol\ specification of \openssl's implementation is limited to the 256-bit digest on input sizes from $0$ bits to $1,087$ bits.
That is a proof that considers all of \figref{fig:sha3}.
The equivalence between \openssl\ \cryptol\ specification and the actual C implementation is limited to the \keccak algorithm only.
That is a proof that only considers \lineref{line:keccak} in \figref{fig:sha3} but holds for any state and is independent of the digest size and rate.

%% TODO: summarize the cost of the total proof in terms of time and identify the machine used for the proof.
